---
title: "Mineral formulae (APFU)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mineral formulae (APFU)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
has_tern <- requireNamespace("ggtern", quietly = TRUE) && requireNamespace("ggarnet", quietly = TRUE)
```

`chemical_formula()` converts oxide wt% (EPMA, SEM-EDS) into atoms per formula unit (APFU),
normalizing either to a fixed number of oxygens (`base = "anions"`) or of cations
(`base = "cations"`).

```{r setup, message = FALSE}
library(ztR)
library(tidyverse)
```

## Input

One row per analysis, an identifier column called `specimen` and one column per oxide, named as
in the oxide table of the package (`SiO2`, `Al2O3`, `FeO`, `Fe2O3`, `H2O_plus`, …). The oxides
must be the last columns of the table, starting at `SiO2`. Missing oxides can be `NA`; they are
treated as zero.

The package ships three small example tables with textbook compositions (Deer, Howie & Zussman):

```{r}
read_ex <- function(file) {
  read.csv(system.file("extdata", file, package = "ztR")) %>%
    select(specimen, mineral, everything(), -X)
}

tourmaline <- read_ex("tourmaline_DHZ.csv")
garnet     <- read_ex("garnet_DHZ.csv")
epidote    <- read_ex("epidote_DHZ.csv")

garnet
```

`chemical_formula()` returns `specimen` plus one column per cation (and the normalization bases),
so the result is joined back to the input by `specimen`.

## Tourmaline: 31 anions

```{r}
tur <- tourmaline %>%
  select(specimen, mineral) %>%
  left_join(chemical_formula(tourmaline, oxigens = 31, cations = 16), by = "specimen")

tur %>% select(mineral, Si, Al, Fe, Fe3, Mg, Na, Ca)
```

## Garnet: 24 oxygens

```{r}
gt <- garnet %>%
  select(specimen, mineral) %>%
  left_join(chemical_formula(garnet, oxigens = 24, cations = 16), by = "specimen")

gt %>% select(mineral, Si, Al, Fe, Fe3, Mn, Mg, Ca)
```

End-member proportions follow directly from the X-site cations:

```{r}
gt %>%
  mutate(x_site = Fe + Mn + Mg + Ca,
         Alm = Fe / x_site, Sps = Mn / x_site, Prp = Mg / x_site, Grs = Ca / x_site) %>%
  select(mineral, Alm:Grs) %>%
  pivot_longer(Alm:Grs, names_to = "end_member", values_to = "fraction") %>%
  ggplot(aes(mineral, fraction, fill = end_member)) +
  geom_col() +
  labs(x = NULL, y = "Molar fraction", fill = NULL) +
  theme_minimal()
```

## Epidote: different oxygen bases per row

`oxigens` can be a vector with one value per row. Here the Fe-free members are normalized to 12.5
oxygens and the rest to 13:

```{r}
epidote %>%
  select(specimen, mineral) %>%
  left_join(chemical_formula(epidote, oxigens = c(12.5, 13, 12.5, 12.5, 12.5, 13), cations = 8),
            by = "specimen") %>%
  select(mineral, Si, Al, Fe3, Ca)
```

## Ternary diagrams

Classification diagrams can be drawn with the companion package
[ggarnet](https://github.com/gabertol/ggarnet), built on ggtern. The chunk below only runs when
both are installed.

```{r, eval = has_tern, message = FALSE}
library(ggtern)
library(ggarnet)

ggarnet_fe_mn() +
  geom_point(data = gt, aes(x = Fe3 + Fe + Mn, z = Ca, y = Mg, shape = mineral), size = 4, alpha = 0.6)

ggarnet_tourmaline() +
  geom_point(data = tur, aes(x = Fe3 + Fe, y = Al, z = Mg, shape = mineral), size = 4, alpha = 0.6)
```
