---
title: "U-Pb geochronology and concordia diagrams"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{U-Pb geochronology and concordia diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
has_isoplotr <- requireNamespace("IsoplotR", quietly = TRUE)
```

ztR computes single-grain U-Pb ages with IsoplotR behind a tidy interface, and draws concordia
diagrams as ordinary ggplot2 layers. This article uses the reference zircons shipped with the
package (91500, Plešovice, FC-1 and PCIGR-1).

The age calculations need IsoplotR (`install.packages("IsoplotR")`); the concordia layers do not.

```{r setup, message = FALSE}
library(ztR)
library(tidyverse)

zircons <- read_csv(system.file("extdata", "stds_zircon.csv", package = "ztR"),
                    show_col_types = FALSE, name_repair = "unique_quiet") %>%
  rename_with(~ .x %>%
                str_replace_all("_mean|_ppm", "") %>%
                str_replace_all("_2se_int", "_2s")) %>%
  mutate(sample = recode(sample, "91500-G" = "91500", "FC1" = "FC-1")) %>%
  mutate(spot = row_number(), .by = sample)

# a manageable subset: 40 analyses per reference material
zr40 <- zircons %>% filter(spot <= 40)
```

## 1. Single-grain ages

`use_isoplotr()` needs the three ratios with their 2σ absolute uncertainties. The error
correlations `rho_206pb_238u_v_207pb_235u` (Wetherill) and `rho_207pb_206pb_v_238u_206pb`
(Tera-Wasserburg) are used when present; otherwise IsoplotR infers them from the redundancy of the
three ratios.

```{r, eval = has_isoplotr}
ages <- use_isoplotr(zr40, legacy = FALSE)

ages %>%
  select(sample, age_68, age_68_2s, age_76, age_76_2s, age_conc, age_conc_2s, p_conc, disc_c) %>%
  head()
```

Every uncertainty column ending in `_2s` is 2σ. `disc_c` is the concordia-distance discordance
(%) of Vermeesch (2021); other definitions are available through `discordance`
(`"r"`, `"t"`, `"a"`, `"sk"`, see `IsoplotR::discfilter()`).

### Speed

The single-grain concordia age is what makes IsoplotR slow (tens of milliseconds per grain). When
only the 7/5, 6/8 and 7/6 ages are needed, `concordia = FALSE` is about 20 times faster:

```{r, eval = has_isoplotr}
all_ages <- use_isoplotr(zircons, concordia = FALSE, legacy = FALSE)
nrow(all_ages)
```

## 2. Comparison with Iolite

The table also carries the ages exported by Iolite (`pb206_u238_age`, `pb207_pb206_age`, …), which
gives a direct check of the calculation:

```{r, eval = has_isoplotr}
all_ages %>%
  transmute(sample,
            `206Pb/238U` = age_68 - pb206_u238_age,
            `207Pb/235U` = age_75 - pb207_u235_age) %>%
  pivot_longer(-sample, names_to = "system", values_to = "difference") %>%
  group_by(sample, system) %>%
  summarise(median_Ma = median(difference, na.rm = TRUE),
            p95_abs_Ma = quantile(abs(difference), 0.95, na.rm = TRUE),
            .groups = "drop")
```

The 206Pb/238U ages agree within a few tenths of a percent (median offset of 3 Ma at 1065 Ma for
91500, 0.2 Ma for Plešovice). The 207Pb/235U ages are systematically older in IsoplotR, by up to
~3% in the low-Pb reference materials. The most likely reason is that Iolite reports the mean of
the ages of each integration, whereas IsoplotR computes the age of the mean ratio; with noisy
207Pb signals the two diverge. The 206Pb/238U uncertainties are identical:

```{r, eval = has_isoplotr}
all_ages %>%
  summarise(ratio_2s = median(age_68_2s / pb206_u238_age_2s, na.rm = TRUE), .by = sample)
```

## 3. Median ages of the reference materials

```{r, eval = has_isoplotr}
ages %>%
  summarise(n = n(),
            concordia_age = median(age_conc, na.rm = TRUE),
            age_68 = median(age_68, na.rm = TRUE),
            .by = sample)
```

Reference values: 91500 ≈ 1065 Ma (Wiedenbeck et al. 1995), Plešovice ≈ 337 Ma (Sláma et al.
2008), FC-1 ≈ 1099 Ma (Paces & Miller 1993).

## 4. Concordia diagrams

`geom_concordia()` draws one error ellipse per row from `x`, `y`, `sigma_x`, `sigma_y` and `rho`.
By default the ellipse is the 95% confidence region; `sigma_level = 2` tells the layer that the
uncertainties mapped to `sigma_x`/`sigma_y` are 2σ. `geom_concordia_line()` adds the concordia
curve with age markers.

### Wetherill

```{r}
zr40 %>%
  filter(sample == "91500") %>%
  ggplot(aes(x = pb207_u235, y = pb206_u238,
             sigma_x = pb207_u235_2s, sigma_y = pb206_u238_2s,
             rho = rho_206pb_238u_v_207pb_235u)) +
  geom_concordia_line(age_range = c(950, 1200), ticks = seq(950, 1200, 50)) +
  geom_concordia(sigma_level = 2, alpha = 0.4, fill = "steelblue", colour = "grey30",
                 linewidth = 0.2) +
  labs(title = "91500", x = "207Pb/235U", y = "206Pb/238U") +
  theme_bw()
```

The concordia line is drawn over `age_range`, which also sets how far the plot extends; choose it
around the ages of the data. Because the line does not know about facets, draw one diagram per
sample (or per age window) rather than faceting samples of very different ages.

### Tera-Wasserburg

Build 238U/206Pb and its uncertainty from 206Pb/238U, and use the Tera-Wasserburg correlation:

```{r}
zr40 %>%
  filter(sample == "PL") %>%
  mutate(u238_pb206 = 1 / pb206_u238,
         u238_pb206_2s = pb206_u238_2s / pb206_u238^2) %>%
  ggplot(aes(x = u238_pb206, y = pb207_pb206,
             sigma_x = u238_pb206_2s, sigma_y = pb207_pb206_2s,
             rho = rho_207pb_206pb_v_238u_206pb)) +
  geom_concordia_line("tw", age_range = c(300, 380), ticks = seq(300, 380, 20)) +
  geom_concordia(sigma_level = 2, alpha = 0.4, fill = "darkorange", colour = "grey30",
                 linewidth = 0.2) +
  labs(title = "Plesovice (PL)", x = "238U/206Pb", y = "207Pb/206Pb") +
  theme_bw()
```

### Colouring by discordance

Any aesthetic can be mapped. Here the ellipses are filled by the concordia-distance discordance,
with IsoplotR's default cutoff (−2% to +5.8%) marked by the outline:

```{r, eval = has_isoplotr}
ages %>%
  filter(sample == "91500") %>%
  mutate(concordant = between(disc_c, -2, 5.8)) %>%
  ggplot(aes(x = pb207_u235, y = pb206_u238,
             sigma_x = pb207_u235_2s, sigma_y = pb206_u238_2s,
             rho = rho_206pb_238u_v_207pb_235u)) +
  geom_concordia_line(age_range = c(900, 1250), ticks = seq(950, 1200, 50)) +
  geom_concordia(aes(fill = disc_c, colour = concordant), sigma_level = 2,
                 alpha = 0.6, linewidth = 0.3) +
  scale_fill_distiller(palette = "RdBu", limits = c(-15, 15), oob = scales::squish) +
  scale_colour_manual(values = c(`TRUE` = "grey20", `FALSE` = "red")) +
  labs(x = "207Pb/235U", y = "206Pb/238U", fill = "disc (%)") +
  theme_bw()
```

## 5. Notes for older code

* `use_isoplotr()` and `tidy_isoplotr()` still return the old columns by default
  (`legacy = TRUE`). **The old `s_2_*` columns hold 1σ values** (they always did); use the
  `*_2s` columns instead. Before version 0.0.0.9001 the data were read in the wrong IsoplotR
  format, so old 207/206 ages, concordia ages and discordances should be recomputed.
* `add_concordia_line()` still works, but its axes are transposed (x = 206Pb/238U,
  y = 207Pb/235U). Use `geom_concordia_line()` for new plots.
* `geom_concordia()` now draws 95% ellipses by default; `level = NULL` reproduces the old 1σ
  ellipses.
