---
title: "Zircon trace-element geochemistry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zircon trace-element geochemistry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

This article walks through the trace-element tools of ztR using the reference zircons shipped
with the package: data quality, chondrite normalization, ratios and proxies, Ti-in-zircon
thermometry, oxybarometry, source-rock classification and parent-rock estimates.

```{r setup, message = FALSE}
library(ztR)
library(tidyverse)

zircons <- read_csv(system.file("extdata", "stds_zircon.csv", package = "ztR"),
                    show_col_types = FALSE, name_repair = "unique_quiet") %>%
  rename_with(~ .x %>%
                str_replace_all("_mean|_ppm", "") %>%
                str_replace_all("_2se_int", "_2s")) %>%
  mutate(sample = recode(sample, "91500-G" = "91500", "FC1" = "FC-1")) %>%
  mutate(spot = row_number(), .by = sample)

ree <- c("la139", "ce140", "pr141", "nd146", "sm147", "eu153", "gd157", "tb159",
         "dy163", "ho165", "er166", "tm169", "yb172", "lu175")
```

## 1. Analytical uncertainty

Before interpreting anything, check how precise each element is. The relative 2σ uncertainty of
the REE per reference material:

```{r, fig.height = 6}
zircons %>%
  select(sample, spot, all_of(ree), all_of(paste0(ree, "_2s"))) %>%
  pivot_longer(-c(sample, spot)) %>%
  mutate(kind = if_else(str_detect(name, "_2s$"), "s2", "value"),
         name = str_remove(name, "_2s$")) %>%
  pivot_wider(names_from = kind, values_from = value) %>%
  mutate(error_pct = 100 * s2 / value,
         element = fct_inorder(str_to_title(str_remove(name, "[0-9]+")))) %>%
  filter(error_pct > 0, error_pct < 200) %>%
  ggplot(aes(element, error_pct)) +
  geom_violin(fill = "grey80", colour = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA) +
  facet_wrap(~ sample) +
  labs(x = NULL, y = "Relative uncertainty (2s, %)") +
  theme_minimal()
```

The LREE (La, Pr) are close to detection limit in zircon and carry very large uncertainties;
ratios using them (e.g. Ce/Ce*) should be read with care.

## 2. Chondrite normalization

`normalize()` divides each element by the CI chondrite of McDonough & Sun (1995) and adds the
normalized columns with the suffix `_N`. Other reference tables can be chosen with `database`
(e.g. `"taylor_mcclennan_1985.csv"` for the upper continental crust).

```{r}
zn <- normalize(zircons, element_vector = c(y89, la139:lu175))
zn %>% select(sample, la139_N:lu175_N) %>% head(3)
```

A spider diagram of the normalized REE:

```{r}
zn %>%
  filter(spot <= 60) %>%
  select(sample, spot, paste0(ree, "_N")) %>%
  pivot_longer(-c(sample, spot), values_to = "value") %>%
  filter(value >= 1e-3) %>%          # drop values at or below detection limit
  mutate(element = fct_inorder(str_to_title(str_remove(name, "[0-9]+_N")))) %>%
  ggplot(aes(element, value, group = spot)) +
  geom_line(alpha = 0.2) +
  scale_y_log10() +
  facet_wrap(~ sample) +
  labs(x = NULL, y = "Sample / CI chondrite") +
  theme_minimal()
```

## 3. Ratios and proxies

`ratio_calculator()` adds the most common zircon ratios in one step. It needs the normalized REE
plus the raw concentrations of Ce, U, Ti, Hf and Y (ppm) and a crystallization age in `best_age`
(Ma), used to correct U for decay in the oxybarometer.

Use `legacy = FALSE` for columns named after what they compute (see `?ratio_calculator`).

```{r}
zr <- zn %>%
  ratio_calculator(legacy = FALSE)

zr %>%
  group_by(sample) %>%
  summarise(across(c(eu_eu, ce_ce, dy_yb, yb_gd, lree_hree, hf_y, fmq, ti_temp),
                   ~ median(.x, na.rm = TRUE)))
```

| Column | Meaning |
|---|---|
| `ce_nd` | Ce_N / Nd_N |
| `dy_yb` | Dy_N / Yb_N |
| `yb_gd` | Yb_N / Gd_N |
| `lree_hree` | (La–Sm)_N / (Gd–Lu)_N |
| `hf_y` | Hf / Y (ppm) |
| `sumREE` | sum of the normalized REE |
| `eu_eu`, `ce_ce` | Eu/Eu\* and Ce/Ce\* (`anomaly()`) |
| `crust` | crustal thickness from Eu/Eu\* (`crustal_thickness()`) |
| `fmq` | ΔFMQ, Loucks et al. (2020) (`FMQ()`) |
| `ti_temp` | Ti-in-zircon temperature, °C (`zircon_ti_t()`) |

## 4. Ti-in-zircon thermometry

`zircon_ti_t()` implements several calibrations. The default `"watson"` is Watson & Harrison
(2005); `"ferry_watson2007"` accounts for the activities of SiO2 and TiO2, which matters when
rutile is absent (aTiO2 < 1).

```{r}
zircons %>%
  filter(ti49 > 0) %>%
  mutate(
    `Watson & Harrison 2005` = zircon_ti_t(ti_ppm = ti49),
    `Ferry & Watson 2007, aTiO2 = 1` = zircon_ti_t(ti_ppm = ti49, equation = "ferry_watson2007"),
    `Ferry & Watson 2007, aTiO2 = 0.6` = zircon_ti_t(ti_ppm = ti49, equation = "ferry_watson2007",
                                                      aTiO2 = 0.6)
  ) %>%
  pivot_longer(starts_with(c("Watson", "Ferry")), names_to = "calibration", values_to = "T_C") %>%
  ggplot(aes(T_C, fill = calibration)) +
  geom_density(alpha = 0.4, colour = NA) +
  facet_wrap(~ sample, scales = "free_y") +
  coord_cartesian(xlim = c(500, 1200)) +
  labs(x = expression(Temperature~(degree*C)), y = NULL, fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

The pressure-dependent calibration of Crisp et al. (2023) is available as `equation = "crisp"`
(with `pressure` in GPa), but its implementation still has to be checked against the paper.

## 5. Source-rock classification

`classify_rock_long()` applies the CART tree of Belousova et al. (2002). It needs Lu, U, Ta and Nb
in ppm, Hf in **wt%**, Ce/Ce\* and Th/U.

```{r}
zr %>%
  mutate(hf = ppm_to_percent(hf177),
         th_u = approx_th / approx_u) %>%
  classify_rock_long() %>%
  count(sample, classification) %>%
  group_by(sample) %>%
  mutate(fraction = n / sum(n)) %>%
  ggplot(aes(sample, fraction, fill = classification)) +
  geom_col() +
  labs(x = NULL, y = "Fraction of analyses", fill = NULL) +
  theme_minimal()
```

**Caveat.** The reference zircons are not all assigned to their known host rocks: Plešovice,
from a potassic granulite, falls mostly in "Carbonatite". The implementation of the tree
(thresholds, units and branch order) is still being checked against Belousova et al. (2002), so
treat these classifications as provisional.

`classify_rock_long_no_ce()` skips Ce/Ce\* (useful when La and Pr are below detection) and
`classify_rock_short()` uses the shorter tree based on Lu, U, Hf, Y and Yb.

## 6. Parent-rock composition and crustal thickness

`WR_calculator()` estimates the trace-element content of the parent melt with the zircon/melt
partition coefficients of Chapman et al. (2016). The result can be normalized again and fed to
the La/Yb crustal-thickness proxy of Profeta et al. (2015):

```{r}
wr <- zircons %>%
  filter(sample == "PL") %>%
  WR_calculator(element_vector = c(y89, nb93, la139:lu175)) %>%
  normalize(tag = "_WR", element_vector = c(y89_WR, nb93_WR, la139_WR:lu175_WR)) %>%
  WR_crustal_thickness()

wr %>%
  summarise(la_yb_N = median(la139_WR_N / yb172_WR_N, na.rm = TRUE),
            thickness_km = median(WR_crust, na.rm = TRUE))
```
