---
title: "Getting started with ztR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ztR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

ztR (**Z**ircon–**T**ourmaline–**R**utile) is a small toolbox for mineral chemistry and zircon
geochronology that works the *tidy* way: a data frame goes in, the same data frame comes out with
new columns. It wraps the heavy lifting of packages such as IsoplotR so you do not have to deal
with their own data classes.

```{r setup, message = FALSE}
library(ztR)
library(tidyverse)
```

## What is in the package

| Topic | Functions | Article |
|---|---|---|
| Mineral formulae (APFU) | `chemical_formula()` | `vignette("mineral-formula")` |
| Zircon trace elements | `normalize()`, `ratio_calculator()`, `anomaly()` | `vignette("zircon-geochemistry")` |
| Thermometry and oxybarometry | `zircon_ti_t()`, `zircon_ti_p()`, `FMQ()` | `vignette("zircon-geochemistry")` |
| Source-rock classification | `classify_rock_long()`, `classify_rock_short()`, `classify_rock_long_no_ce()` | `vignette("zircon-geochemistry")` |
| Parent-rock composition | `WR_calculator()`, `WR_crustal_thickness()`, `crustal_thickness()` | `vignette("zircon-geochemistry")` |
| U-Pb ages | `use_isoplotr()` | `vignette("u-pb-geochronology")` |
| Concordia diagrams | `geom_concordia()`, `geom_concordia_line()` | `vignette("u-pb-geochronology")` |

## Conventions

ztR expects lower-case column names made of the element symbol and the mass of the measured
isotope, and uncertainties in a column with the same name plus the suffix `_2s`:

| Quantity | Column | Uncertainty |
|---|---|---|
| La (ppm) | `la139` | `la139_2s` |
| U (ppm) | `approx_u` | `approx_u_2s` |
| 207Pb/235U | `pb207_u235` | `pb207_u235_2s` |
| 206Pb/238U | `pb206_u238` | `pb206_u238_2s` |
| 207Pb/206Pb | `pb207_pb206` | `pb207_pb206_2s` |
| Wetherill error correlation | `rho_206pb_238u_v_207pb_235u` | — |
| Tera-Wasserburg error correlation | `rho_207pb_206pb_v_238u_206pb` | — |

* Concentrations are in ppm.
* Uncertainties are **2-sigma absolute**.
* Ages are in Ma.
* Chondrite-normalized values get the suffix `_N`; whole-rock estimates the suffix `_WR`.

## Reading an Iolite export

The package ships a table of reference zircons (91500, Plešovice, FC-1 and PCIGR-1) exported from
Iolite. Iolite names columns like `la139_ppm_mean` and `la139_ppm_2se_int`; one `rename_with()`
brings them to the ztR convention:

```{r}
zircons <- read_csv(system.file("extdata", "stds_zircon.csv", package = "ztR"),
                    show_col_types = FALSE, name_repair = "unique_quiet") %>%
  rename_with(~ .x %>%
                str_replace_all("_mean|_ppm", "") %>%
                str_replace_all("_2se_int", "_2s")) %>%
  mutate(sample = recode(sample, "91500-G" = "91500", "FC1" = "FC-1"))

zircons %>% count(sample)
```

The same object is used in the other articles.

## A first pipeline

Every function adds columns, so they chain:

```{r}
zircons %>%
  filter(sample == "PL") %>%
  slice(1:5) %>%
  normalize(element_vector = c(la139:lu175)) %>%
  mutate(eu_eu = anomaly(eu153_N, sm147_N, gd157_N),
         t_ti  = zircon_ti_t(ti_ppm = ti49)) %>%
  select(sample, eu_eu, t_ti)
```

## Backwards compatibility

Version 0.0.0.9001 fixed several bugs (see `NEWS.md`). Old code keeps running:

* `ratio_calculator()` and `use_isoplotr()` return their old columns by default
  (`legacy = TRUE`) and warn once per session about what those columns really contain;
* `add_concordia_line()` still exists, with its original (transposed) axes;
  `geom_concordia_line()` is the new version.
